Better fusion for filters

### 2014-03-27 2014betterfusionforfilters/section/04-ILP.tex
add fusion preventing constraint
mention quadtree
start filling in ILP formulation and weights

ILP variables:
\begin{tabbing}
M   \= M \= MMMMMMM \= MM \= \kill
$x$ \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$ \\
$\pi$ \> @::@  \> $node$ \> $\to$ \> $\mathbb{N}$ \\
\end{tabbing}

@@

Use the lemma:
a graph $G$ is acyclic if and only if there exists some mapping $\pi :: node \to \mathbb{N}$ such that for every edge in $G$ from $i$ to $j$, $\pi_i < \pi_j$.

\begin{tabbing}
MMMMM   \= MMMMMMMMMMM \= \kill
Minimise   \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$       \> ($W_{ij}$ is constant weight)      \\
Subject to \> $-N x_{ij} \le \pi_j - \pi_i \le N x_{ij}$ \> (no edge from i to j)            \\
           \> $              \pi_j - \pi_i \le N x_{ij}$ \> (an edge from i to j)            \\
\end{tabbing}

@@

@@

For a clustering to be legal, it must satisfy three constraints:
\begin{description}
\item[Acyclic]
after merging nodes of same cluster together, the resulting graph must be a dag
\item[Precedence preserving]
if there is an edge between two nodes $i$ and $j$, and they are not merged together, then we require $\pi_j > \pi_i$
\item[Type constraint]
if two nodes $i$ and $j$ are in the same cluster, then $\tau_i = \tau_j$, or if $\tau_i$ is a subtype of $\tau_j$ (or $\tau_j$ is a subtype of $\tau_i$), then the \emph{generator} for $\tau_i$ (or $\tau_j$) must also be in the same cluster as $i$ and $j$.
    \\
    \TODO Actually, let us say $x_{ij} = 0 \implies check_{ij}$

>>>>

ILP variables:
\begin{tabbing}
M   \= M \= MMMMMMM \= MM \= \kill
$x$   \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$ \\
      \>       \> ($x_{ij} = 0$ iff $i$ and $j$ are fused together) \\
$\pi$ \> @::@  \> $node$             \> $\to$ \> $\mathbb{R}$ \\
      \>       \> ($\pi_i$ is used to ensure an acyclic clustering) \\
$c$   \> @::@  \> $node$             \> $\to$ \> $\mathbb{B}$ \\
      \>       \> ($c_i = 0$ iff $\forall (i,j) \in E. x_{ij} = 0$) \\
      \>       \> (i.e. no intermediate array necessary)            \\
\end{tabbing}

@@

Use the lemma:
a graph $G$ is acyclic if and only if there exists some mapping $\pi :: node \to \mathbb{R}$ such that for every edge in $G$ from $i$ to $j$, $\pi_i < \pi_j$.

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \> (no edge from i to j)            \\
           \>             \>       \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \> (an edge from i to j)            \\
\\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \> (for all edges from i to j)      \\
\\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \> (fusion-preventing edges from i to j)      \\
\\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \> (if $\tau_i \not= \tau_j \wedge gen_i=\emptyset \wedge gen_j=\emptyset$)                 \\
           \> $x_{ij}$    \> $\le$ \> $x_{i'j}$       \>       \>            \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
           \> $x_{ij}$    \> $\le$ \> $x_{ij'}$       \>       \>            \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
\end{tabbing}

\subsubsection{The benefits of fusion}

Fusing loops can have three main benefits:
\begin{itemize}
\item
reducing memory traffic, such as multiple loops reading from the same array;
\item
removing intermediate arrays, thus reducing the amount of memory required;
\item
and reducing loop overhead, such as when two loops operate on different arrays of the same size.
\end{itemize}
However, these benefits cannot be considered in isolation; for example, fusing two loops to reduce loop overhead may remove potential fusion opportunities that reduce memory traffic.
When operating on large arrays that do not fit in cache, memory traffic dominates execution time.
An excessive number of intermediate arrays can also cause issues if all are live in memory at once, potentially leading to \emph{thrashing}.
The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion.

This total ordering can be encoded into an ILP objective function as weights.
If the program graph contains $n$ combinators, then there are at most $n$ opportunities for fusion.
The encoding of loop overhead is weight $1$, removing intermediate arrays is weight $n$, and reducing memory traffic is weight $n^2$.
This ensures that no amount of loop overhead reduction can outweigh the benefit of removing an intermediate array,
and likewise no number of removed intermediate arrays can outweigh a reduction in memory traffic.
Although, it is worth noting that reducing memory traffic does \emph{tend} to remove intermediate arrays, and vice versa.

\begin{tabbing}
MMMMM   \= MMMM \= M \= \kill
Minimise   \> \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
Where      \>                                       \\
           \> $W_{ij} = N^2$ \> $|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{ij} = N^2$ \> $|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{ij} = 1$   \> $|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        \\
           \\
           \> $N = |V|$
\end{tabbing}

@@

\begin{tabbing}
MMMMMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
$split$     \> @::@ \> $V \times E \to \{ \{name\} \}$      \\
MMMM        \= M    \= \kill
$split(vs,es)$ \\
    \> $=$  \> $\{clusterable(v,vs,es) | v \in vs\}$      \\
$clusterable(v,vs,es)$  \\
    \>$=$       \>$\{v' | v' \in vs$                                \\
    \>$\wedge$  \>$\forall p. p = path(v,v') \vee p = path(v',v)$   \\
    \>$\wedge$  \>$fusion~preventing \not\in p\}$\\
\end{tabbing}

@@

For a clustering to be legal, it must satisfy three constraints:
\begin{description}
\item[Acyclic]
after merging nodes of same cluster together, the resulting graph must be a dag
\item[Precedence preserving]
if there is an edge between two nodes $i$ and $j$, and they are not merged together, then we require $\pi_j > \pi_i$
\item[Fusion preventing]
likewise, if there is a fusion-preventing edge between two nodes $i$ and $j$, then we require $\pi_j > \pi_i$, which implies that they are not merged together
\item[Type constraint]
if two nodes $i$ and $j$ are in the same cluster, then $\tau_i = \tau_j$, or if $\tau_i$ is a subtype of $\tau_j$ (or $\tau_j$ is a subtype of $\tau_i$), then the \emph{generator} for $\tau_i$ (or $\tau_j$) must also be in the same cluster as $i$ and $j$.
    \\
    \TODO Actually, let us say $x_{ij} = 0 \implies check_{ij}$
